---
title: "Wrangling - tidyr"
author: "Elise Gonzales"
date: "2022-08-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)

library(tidyverse)
library(here)
library(janitor)
```

## Read in WB data files.
```{r}
wb_indicators <- read_csv(here("data", "wb_indicators.csv"), na = c("..", "")) #converts .. to true "NA"
wb_metadata <- read_csv(here("data", "wb_indicators_metadata.csv"))
```

##Data wrangling
###pivot_longer() to combine years
```{r}
wb_indicators_long <- wb_indicators |> 
  pivot_longer(cols = `2001 [YR2001]`:`2020 [YR2020]`,
               names_to = "year",
               values_to = "indicator_value")

class(wb_indicators_long$year)
```

###separate() to clean up year column & exclude some cols
```{r}
#check output after every step!
class(wb_clean$year)
#year still a character. Pipe above and use as.numeric()

wb_clean <- wb_indicators_long |> 
  separate(col = year,
           into = c("year", "year_chr"),
           sep = " ") |> 
  select(-year_chr, -`Country Code`, -`Series Code`) |> #use to exclude some cols 
  mutate(year = as.numeric(year)) |> 
  drop_na(`Series Name`) |> 
  pivot_wider(names_from = `Series Name`, values_from = indicator_value)

names(wb_clean) <- c("country", "year", "access_clean_fuels_pp", "access_electricity_pp", "co2_emissions_kt", "fossil_fuel_cons_pct", "water_stress")
  
wb_subset <- wb_clean |> 
  filter(country %in% c("Algeria", "Botswana", "China", "United States"))

unique(wb_subset$country)
```

###ggplot
```{r}
ggplot(data = wb_subset, aes(x = year, y = co2_emissions_kt)) + geom_line(aes(color = country))  + facet_wrap(~country)
```

##Filtering rows
Example 1: filter to only keep exact matches
```{r}
ex_1 <- starwars |> 
  filter(height > 180)
```

```{r}
ex_2 <- starwars |> 
  filter(eye_color == "blue")
#run unique() in console to check if this filter worked
```

```{r}
ex_3 <- starwars |> 
  filter(homeworld == "Naboo")
```

```{r}
ex_4 <- starwars |> 
  filter(height > 180 & homeworld == "Tatooine")
```

```{r}
ex_5 <- starwars |> 
  filter(hair_color == "brown" & species == "Human")
```

###filtering OR statements
```{r}
ex_6 <- starwars |> 
  filter(height > 180 | eye_color == "yellow")
```

```{r}
ex_7 <- starwars |> 
  filter(homeworld == "Endor" | species == "Droid")
```

```{r}
ex_8 <- starwars |> 
  filter(eye_color == "blue" | eye_color == "brown" | eye_color == "red")
#better way to do this below:
ex_9 <- starwars |> 
  filter(eye_color %in% c("blue", "brown", "red")) # Does the value in this row match blue or brown or red. Then goes to next line and asks the same question. "==" would apply question: does row one match blue, does row 2 match brown, does row 3 match red. and it'll keep recycling. instead of asking for each color for EACH row.
#NEVER DO THIS
ex_10 <- starwars |> 
  filter(eye_color == c("blue", "brown", "red")) # "==" is to look for SINGLE, EXACT MATCH
```

```{r}
ex_11 <- starwars |> 
  filter(homeworld %in% c("Endor", "Tatooine", "Naboo"), height < 150) #can use "," or "and"
```

###Negate filter
```{r}
ex_12 <- starwars |> 
  filter(species != "Human")

ex_13 <- starwars |> 
  filter(!species %in% c("Human", "Ewok", "Wookiee"))
```

##Select/exclude cols
```{r}
#Select by name
ex_14 <- wb_clean |> 
  select(country, year, co2_emissions_kt)
ex_15 <- wb_clean |> 
  select(year:fossil_fuel_cons_pct)

#exclude one col
ex_16 <- wb_clean |> 
  select(year:fossil_fuel_cons_pct, -access_electricity_pp)
```

```{r}
#pipe through mult sequences. Check along the way!
ex_17 <- wb_clean |> 
  filter(country %in% c("Bolivia", "Chile", "Mexico")) |> 
  select(country:fossil_fuel_cons_pct)
```

```{r}
ex_18 <- wb_clean |> 
  select(year_new = year,
         emissions = co2_emissions_kt,
         country_name = country)
```

```{r}
ex_19 <- wb_clean |> 
  select(country, year, water_stress) |> 
  filter(country %in% c("China", "Australia", "Canada"))
```

###rename
```{r}
ex_20 <- wb_clean |> 
  rename(ffcp = fossil_fuel_cons_pct,
         ws = water_stress)

ex_21 <- ex_20 |> #new name = old name
  rename(access_clean = access_clean_fuels_pp,
         access_elec = access_electricity_pp)
```

###mutate: update existing col/add new col

If you assign a NEW col name, it will add a new col to the end of your data frame. If you assign an existing col name, it will overwrite that col.
```{r}
ex_22 <- wb_clean |> #add new col & relocate
  mutate(co2_emissions_t = co2_emissions_kt * 1000) |> 
  relocate(co2_emissions_t, .after = co2_emissions_kt)

ex_23 <- wb_clean |> 
  mutate(yay = "YAY")

ex_24 <- wb_clean |> 
  mutate(year = as.character(year))

ex_25 <- ex_24 |> 
  mutate(year = as.numeric(year))

# dplyr::relocate()
ex_26 <- wb_clean |> 
  relocate(country, .after = year)
```

### `group_by() |> summarize()`
```{r}
#group_by(): creates 
#summarize(): apply your funcs to EACH GROUP, and return them in a summary table
ex_27 <- starwars |> 
  filter(homeworld %in% c("Naboo", "Tabooine")) |> 
  group_by(species, homeworld) |> 
  summarize(mean_height = mean(height, na.rm = TRUE),
            mean_mass = mean(mass, na.rm = TRUE))
```

##Frequency/Case table
```{r}
df <- tribble(
  ~species, ~length, ~number,
  "lobster", 12, 4,
  "lobster", 14, 6,
  "lobster", 15, 2
)

df_case <- df |> 
  uncount(number)

#opposite 
df_count <- df_case |> 
  dplyr::count(length)
#example -- count does --> group_by(), summarize(), n()
sw_count <- starwars |> 
  count(homeworld, species)
```

##Practice example
Start from wb_clean. 
- Only keep obs for Nicaragua, Ecuador, Peru, & Chile. 
- For years > 2005. 
- Keep columns for country, year, & co2 emissions. 
- Add a new col that is co2_emissions converted to megatons.
- Find the total co2 emissions in megatons for years still in the dataset by country, using group by/summarize.
```{r}
wb_practice <- wb_clean |> 
  filter(country %in% c("Nicaragua", "Ecuador", "Peru", "Chile"), year > 2005) |>
  select(country, year, co2_emissions_kt) |> 
  mutate(co2_emissions_mt = co2_emissions_kt * 0.001) 
wb_prac_cont <- wb_practice |> 
  group_by(country) |> 
  summarize(tot_co2_emissions_mt = mean(co2_emissions_mt, na.rm = TRUE))
  
```

